Recent advances in random matrix theory applied to heavy-tailed multivariate time series

### 2015-09-09 papers/Number1/paper.tex
Revised the paper.

Figure \ref{fig:LamYao} compares the largest eigenvalue of the sum of
time-lagged covariance matrices with the sum of the largest
eigenvalues of individual time-lagged covariance matrices. The data
are rank-transformed return series of 478 stocks selected from the S\&P
500 index.
\begin{figure}[htb!]
  \centering
  \subfigure[] {
    \includegraphics[scale=0.4]{eigen_sum.pdf}
    \label{fig:LamYao:a}
  }
  \subfigure[] {
    \includegraphics[scale=0.4]{eigen_sum_plus.pdf}
    \label{fig:LamYao:a}
  }
  \label{fig:LamYao}
  \caption{The largest eigenvalue of summed time-lagged covariance
    matrices and the sum of the largest eigenvalues of time-lagged
    covariance matrices. In both cases $(C_i)_{lm} =
    (np)^{-2}\sum_{t=1}^{n-i} X_{lt} X_{m,t+i}$}
\end{figure}

\subsection{Overview of light-tailed
  case}\label{sec:light_tail_review}
As early as in 1962, T. W. Anderson \cite{anderson:1963} showed that,
when the entries of the data matrix has normal distribution ($\E
X_{ij} = 0$ for $i,j = 1,2,\dots, p$; $\E X_{ii}^2 = 2\sigma^2, \E
X_{ij}^2 = \sigma^2$ for $i \neq j$), the joint density function
of the eigenvalues of the sample covariance matrix is given by
\[
f(\lambda_1, \lambda_2, \dots, \lambda_p; \sigma, p) \propto \prod_{i=1}^p \exp\left(
  -{\lambda_i^2 \over 4\sigma^2}
\right) \prod_{i < j} (\lambda_i - \lambda_j)
\]
In 1988, Yin, Bai and Krishnaiah proved in
\cite{yin:bai:Krishnaiah:1988} the following result for a $p \times n$
matrix $\bf X$ with iid entries $X_{ij}$ under the assumption $\E
X_{11} = 0$, $\E X_{11}^4 < \infty$, and $q = \lim_{n \to \infty} p/n
> 0$
\[
\lambda_{\M}\left({1 \over n}\bf{XX'}\right) \overset{\as}{\to} \left(
  1 + \sqrt{q}
\right)^2 \E X_{11}^2 \text{ as }n \to \infty
\]
When $\E X_{11}^4 = \infty$, Yin and Silverstein proved in 1988
\cite{yin:silverstein:1988}
\[
\lim_{n \to \infty} \sup \lambda_{\M}\left({1 \over
    n}\bf{XX'}\right) = \infty
\]
Later in 2001, I. M. Johnstone showed that, when the data matrix has
standard normal distribution, i.e. $X_{ij} \sim N(0, 1)$, the largest
eigenvalue of the sample covariance matrix follows the Tracy-Widom
distribution:
\[
{ł_1 - \mu_{np} \over \sigma_{np}} \overset{d}{\to} W_1 \sim F_1
\text{ as }{n, p \to \infty}
\]
where $\mu_{np}$ and $\sigma_{np}$ are constants dependent on $p$ and
$n$ and the distribution function $F_1$ is given by
\[
F_1(s) = \exp\left\{
  -{1 \over 2} \int_{s}^\infty \left[
    q(x) + (x - s) q^2(x)
  \right] dx
\right\}
\]
where $q(x)$ is the unique solution to the Painlev\'e II differential
equation
\begin{eqnarray*}
  q''(x) &=& xq(x) + 2 q^3(x) \\
  q(x) &\sim& \text{Ai}(x) \text{ as } x \to \infty
\end{eqnarray*}
Figure \ref{fig:normal-3point-TW} compares the density function of the largest
eigenvalue from simulated $N(0,1)$ and 3-point samples with
that of the Tracy-Widom.
\begin{figure}[htb!]
  \centering
  \subfigure[$N(0, 1)$ entries]{
    \includegraphics[scale=0.4]{normal-TW.pdf}
  }
  \subfigure[Entry distribution: $\P(X=\sqrt 3) = \P(X=-\sqrt 3 ) =
  1/6$, $\P(X=0)=2/3$]{
    \includegraphics[scale=0.4]{3point-TW.pdf}
  }
  \caption{Sample density function of the largest eigenvalue compared
    with the Tracy-Widom density function. The data matrix has
    dimension $200 \times 1000$. An ensemble of 2000 matrices are
    simulated.}
  \label{fig:normal-3point-TW}
\end{figure}
In the case of entry distributions with infinite 4th moment, the
Tracy-Widom law no longer applies and gives way to the Frechet
distribution. Figure \ref{fig:MyDist-Frechet} illustrates this with a
simulated ensemble whose entries are distributed according to
\eqref{eq:mydist}.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5]{MyDist-Frechet.pdf}  
  \caption{Sample distribution function of the largest eigenvalue compared with that of Frechet}
  \label{fig:MyDist-Frechet}
\end{figure}
Later on, several authors \cite{Pillai:Yin:2011, tao:vu:2012,
  Wang:2012} looked into sample covariance matrix $S_{n, p} = {1
  \over n} \bf{XX}'$, where the entries of $\bf X$, denoted $X_{ij}$
are independent and satisfy the following condition which is referred
to as condition {\bf C0}.
\begin{enumerate}
\item $\E X_{ij} = 0$, $\E X_{ij}^2 = 1$ and
\item $\forall u > 1, \exists \theta > 0$ such that
  \[
  \P(X_{ij} > u) = \theta^{-1} e^{-u^{\theta}}
  \]
\end{enumerate}
Under condition {\bf C0}, Pillai and Yin \cite{Pillai:Yin:2011} proved
that, for $q = \lim_{n \to \infty} {p \over n} \in (0, \infty)
\setminus \{1\}$, the probability of the $j$-th largest eigenvalue of
the sample covariance matrix deviating significantly from its
theoretical value computed as the $1 - j/p$ quantile of the
Marcenko-Pastur distribution decreases exponentially\footnotemark.

>>>>

\newpage
Figure \ref{fig:LamYao} compares the largest eigenvalue of the sum of
time-lagged covariance matrices with the sum of the largest
eigenvalues of individual time-lagged covariance matrices. The data
are return series of 478 stocks selected from the S\&P 500 index.
\begin{figure}[htb!]
  \centering
  \subfigure[] {
    \includegraphics[scale=0.4]{eigen_sum.pdf}
    \label{fig:LamYao:a}
  }
  \subfigure[] {
    \includegraphics[scale=0.4]{eigen_sum_plus.pdf}
    \label{fig:LamYao:b}
  }
  \caption{The largest eigenvalue of summed time-lagged covariance
    matrices and the sum of the largest eigenvalues of time-lagged
    covariance matrices. In both cases $(C_i)_{lm} =
    \sum_{t=1}^{n-i} X_{lt} X_{m,t+i}$.}
  \label{fig:LamYao}
\end{figure}

\newpage

\subsection{Overview of light-tailed
  case}\label{sec:light_tail_review}
As early as in 1962, T. W. Anderson \cite{anderson:1963} showed that,
when the entries of the data matrix has normal distribution ($\E
X_{ij} = 0$ for $i,j = 1,2,\dots, p$; $\E X_{ii}^2 = 2\sigma^2, \E
X_{ij}^2 = \sigma^2$ for $i \neq j$), the joint density function
of the eigenvalues of the sample covariance matrix is given by
\[
f(\lambda_1, \lambda_2, \dots, \lambda_p; \sigma, p) \propto \prod_{i=1}^p \exp\left(
  -{\lambda_i^2 \over 4\sigma^2}
\right) \prod_{i < j} (\lambda_i - \lambda_j)
\]
In 1988, Yin, Bai and Krishnaiah proved in
\cite{yin:bai:Krishnaiah:1988} the following result for a $p \times n$
matrix $\bf X$ with iid entries $X_{ij}$ under the assumption $\E
X_{11} = 0$, $\E X_{11}^4 < \infty$, and $q = \lim_{n \to \infty} p/n
> 0$
\[
\lambda_{\M}\left({1 \over n}\bf{XX'}\right) \overset{\as}{\to} \left(
  1 + \sqrt{q}
\right)^2 \E X_{11}^2 \text{ as }n \to \infty
\]
When $\E X_{11}^4 = \infty$, Yin and Silverstein proved in 1988
\cite{yin:silverstein:1988}
\[
\lim_{n \to \infty} \sup \lambda_{\M}\left({1 \over
    n}\bf{XX'}\right) = \infty
\]
Later in 2001, I. M. Johnstone showed that, when the data matrix has
standard normal distribution, i.e. $X_{ij} \sim N(0, 1)$, the largest
eigenvalue of the sample covariance matrix follows the Tracy-Widom
distribution:
\[
{ł_1 - \mu_{np} \over \sigma_{np}} \overset{d}{\to} W_1 \sim F_1
\text{ as }{n, p \to \infty}
\]
where $\mu_{np}$ and $\sigma_{np}$ are constants dependent on $p$ and
$n$ and the distribution function $F_1$ is given by
\[
F_1(s) = \exp\left\{
  -{1 \over 2} \int_{s}^\infty \left[
    q(x) + (x - s) q^2(x)
  \right] dx
\right\}
\]
where $q(x)$ is the unique solution to the Painlev\'e II differential
equation
\begin{eqnarray*}
  q''(x) &=& xq(x) + 2 q^3(x) \\
  q(x) &\sim& \text{Ai}(x) \text{ as } x \to \infty
\end{eqnarray*}
Figure \ref{fig:normal-3point-TW} compares the density function of the largest
eigenvalue from simulated $N(0,1)$ and 3-point samples with
that of the Tracy-Widom.
\begin{figure}[htb!]
  \centering
  \subfigure[$N(0, 1)$ entries]{
    \includegraphics[scale=0.4]{normal-TW.pdf}
  }
  \subfigure[Entry distribution: $\P(X=\sqrt 3) = \P(X=-\sqrt 3 ) =
  1/6$, $\P(X=0)=2/3$. Note $\E X = 0$, $\E X^2 = 1$, $\E X^3 = 0$ and
  $\E X^4 = 3$, i.e. the first 4 moments match those of $N(0, 1).$]{
    \includegraphics[scale=0.4]{3point-TW.pdf}
  }
  \caption{Sample density function of the largest eigenvalue compared
    with the Tracy-Widom density function. The data matrix has
    dimension $200 \times 1000$. An ensemble of 2000 matrices are
    simulated.}
  \label{fig:normal-3point-TW}
\end{figure}

\newpage
In the case of entry distributions with infinite 4th moment, the
Tracy-Widom law no longer applies and gives way to the Frechet
distribution. Figure \ref{fig:MyDist-Frechet} illustrates this with a
simulated ensemble whose entries are distributed according to
\eqref{eq:mydist} with $\alpha = 1.6$.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5]{MyDist-Frechet.pdf}  
  \caption{Sample distribution function of the largest eigenvalue
    compared with that of Frechet. The data matrices have dimensions
    $200 \times 1000$. An ensemble of 2000 matrices are simulated.}
  \label{fig:MyDist-Frechet}
\end{figure}

\newpage
Later on, several authors \cite{Pillai:Yin:2011, tao:vu:2012,
  Wang:2012} looked into sample covariance matrix $S_{n, p} = {1
  \over n} \bf{XX}'$, where the entries of $\bf X$, denoted $X_{ij}$
are independent and satisfy the following condition which is referred
to as condition {\bf C0}.
\begin{enumerate}
\item $\E X_{ij} = 0$, $\E X_{ij}^2 = 1$ and
\item $\forall u > 1, \exists \theta > 0$ such that
  \[
  \P(X_{ij} > u) = \theta^{-1} e^{-u^{\theta}}
  \]
\end{enumerate}
Under condition {\bf C0}, Pillai and Yin \cite{Pillai:Yin:2011} proved
that, for $q = \lim_{n \to \infty} {p \over n} \in (0, \infty)
\setminus \{1\}$, the probability of the $j$-th largest eigenvalue of
the sample covariance matrix deviating significantly from its
theoretical value computed as the $1 - j/p$ quantile of the
Marcenko-Pastur distribution decreases exponentially\footnotemark.
